---
title: "honey-dashboard"
format:
  dashboard:
    css: honey-page.css
    mainfont: "Fira Code"
execute:
  echo: false
  warning: false
  message: false
---

::: {.page-nav}

[Overview](danl310proj.html){.btn .btn-outline-primary}
[Dashboard](honey-dashboard.html){.btn .btn-primary}

:::

```{r}
library(tidyverse)
library(maps)
library(ggthemes)
library(scales)
library(ggrepel)

honey_gold <- "#D88400"
honey_yellow <- "#FFC400"
honey_dark <- "#3A2500"
honey_light <- "#FFF4CC"

theme_honey_light <- function(base_size = 14) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid.major = element_line(color = "#E8D7A3", linewidth = 0.3),
      panel.grid.minor = element_blank(),
      plot.title = element_text(
        face = "bold",
        color = honey_dark,
        size = 17,
        lineheight = 1.05
      ),
      plot.subtitle = element_text(
        color = "#6A4A00",
        size = 11,
        lineheight = 1.05
      ),
      plot.caption = element_text(color = "#7A6500", size = 9, hjust = 0),
      axis.title = element_text(face = "bold", color = honey_dark),
      axis.text = element_text(color = honey_dark),
      legend.title = element_text(face = "bold", color = honey_dark),
      legend.text = element_text(color = honey_dark),
      strip.background = element_rect(fill = honey_light, color = NA),
      strip.text = element_text(face = "bold", color = honey_dark),
      legend.position = "right",
      plot.margin = margin(12, 20, 12, 20)
    )
}

honey_data <- read_csv("honeyproduction.csv")
```

::: {.dashboard-intro}

## Honey Production Dashboard

This dashboard visualizes trends in U.S. honey production from 1998 to 2012. The plots focus on geographic production patterns, national production change, state-level declines, price relationships, and the importance of top-producing states.


## Dataset Summary

```{r}
#| output: asis

honey_summary_table <- honey_data %>%
  summarize(
    `Number of rows` = n(),
    `Number of states` = n_distinct(state),
    `First year` = min(year, na.rm = TRUE),
    `Last year` = max(year, na.rm = TRUE),
    `Average colonies` = mean(numcol, na.rm = TRUE),
    `Average yield per colony` = mean(yieldpercol, na.rm = TRUE),
    `Average total production` = mean(totalprod, na.rm = TRUE),
    `Average price per lb` = mean(priceperlb, na.rm = TRUE),
    `Average production value` = mean(prodvalue, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Dataset Feature",
    values_to = "Value"
  ) %>%
  mutate(
    Value = case_when(
      `Dataset Feature` %in% c(
        "Average colonies",
        "Average yield per colony",
        "Average total production",
        "Average price per lb",
        "Average production value"
      ) ~ comma(round(as.numeric(Value), 2)),
      TRUE ~ as.character(Value)
    )
  )

rmarkdown::paged_table(honey_data,
                       options = list(rows.print = 20,
                                      cols.print = 6,
                                      pages.print = 0,
                                      paged.print = F
                                      )) 
```


## Average Honey Production by State

```{r}
#| fig-width: 11
#| fig-height: 6
#| out-width: "100%"
#| fig-align: center

df <- honey_data %>% 
  select(state, year, totalprod)

state_lookup <- tibble(
  abb = state.abb,
  region = str_to_lower(state.name)
)

df2 <- df %>%
  mutate(state = str_trim(str_to_upper(state))) %>%
  left_join(state_lookup, by = c("state" = "abb")) %>%
  group_by(region) %>%
  summarize(
    avg_totalprod = mean(totalprod, na.rm = TRUE),
    .groups = "drop"
  )

us_states <- map_data("state")

map_df <- us_states %>%
  left_join(df2, by = "region")

p_map <- ggplot(
  map_df,
  aes(x = long, y = lat, group = group, fill = avg_totalprod)
) +
  geom_polygon(color = "white", linewidth = 0.25) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  scale_fill_gradient(
    low = "#FFF8DC",
    high = honey_gold,
    name = "Average honey\nproduction",
    na.value = "gray92",
    labels = label_number(scale_cut = cut_short_scale())
  ) +
  labs(
    title = str_wrap("Average Honey Production by State", width = 45),
    subtitle = str_wrap("Average total production across the full 1998–2012 timeframe", width = 65),
    caption = "Data: honeyproduction.csv"
  ) +
  theme_map() +
  theme_honey_light() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "right"
  )

p_map
```


## U.S. Honey Production Trends, 1998–2012

```{r}
#| fig-width: 11
#| fig-height: 7
#| out-width: "100%"
#| fig-align: center

us_year <- honey_data %>%
  group_by(year) %>%
  summarize(
    totalprod = sum(totalprod, na.rm = TRUE),
    numcol = sum(numcol, na.rm = TRUE),
    yieldpercol = mean(yieldpercol, na.rm = TRUE),
    priceperlb = mean(priceperlb, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(totalprod, numcol, yieldpercol, priceperlb),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(
      metric,
      levels = c("totalprod", "numcol", "yieldpercol", "priceperlb"),
      labels = c(
        "Total production (lbs)",
        "Number of colonies",
        "Yield per colony (lbs)",
        "Price per lb ($)"
      )
    )
  )

p1 <- ggplot(us_year, aes(x = year, y = value)) +
  geom_point(size = 2.2, color = honey_gold) +
  geom_line(linewidth = 1, color = honey_dark) +
  geom_smooth(
    show.legend = FALSE,
    color = honey_gold,
    fill = honey_light,
    linewidth = 0.9
  ) +
  facet_wrap(~metric, scales = "free_y", ncol = 2) +
  scale_x_continuous(breaks = seq(1998, 2012, by = 2)) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(
    x = NULL,
    y = NULL,
    title = str_wrap("U.S. Honey Production Trends, 1998–2012", width = 45),
    subtitle = str_wrap("Production, colonies, yield, and price move in different directions over time", width = 70),
    caption = "Data: honeyproduction.csv"
  ) +
  theme_honey_light()

p1
```



## Largest Production Declines

```{r}
#| fig-width: 10
#| fig-height: 6
#| out-width: "100%"
#| fig-align: center

state_change <- honey_data %>%
  filter(year %in% c(1998, 2012)) %>%
  select(state, year, totalprod) %>%
  pivot_wider(
    names_from = year,
    values_from = totalprod,
    names_prefix = "year_"
  ) %>%
  mutate(change = year_2012 - year_1998) %>%
  slice_min(change, n = 10) %>%
  arrange(change)

state_change_long <- state_change %>%
  pivot_longer(
    cols = c(year_1998, year_2012),
    names_to = "year",
    values_to = "totalprod"
  ) %>%
  mutate(
    year = factor(
      year,
      levels = c("year_1998", "year_2012"),
      labels = c("1998", "2012")
    ),
    state = fct_inorder(state)
  )

p2 <- ggplot(
  state_change_long,
  aes(x = year, y = totalprod, group = state)
) +
  geom_line(linewidth = 1.2, color = honey_gold) +
  geom_point(size = 3.4, color = honey_dark) +
  geom_text(
    data = state_change_long %>% filter(year == "1998"),
    aes(label = state),
    hjust = 1.2,
    size = 3.5,
    color = honey_dark,
    show.legend = FALSE
  ) +
  geom_text(
    data = state_change_long %>% filter(year == "2012"),
    aes(label = state),
    hjust = -0.2,
    size = 3.5,
    color = honey_dark,
    show.legend = FALSE
  ) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  coord_cartesian(clip = "off") +
  labs(
    x = NULL,
    y = "Total production (lbs)",
    title = str_wrap("Largest Production Declines", width = 45),
    subtitle = str_wrap("Comparing state production in 1998 and 2012", width = 65),
    caption = "Data: honeyproduction.csv"
  ) +
  theme_honey_light() +
  theme(
    plot.margin = margin(10, 60, 10, 60),
    legend.position = "none"
  )

p2
```



## Annual U.S. Production Compared with Honey Price

```{r}
#| fig-width: 10
#| fig-height: 6
#| out-width: "100%"
#| fig-align: center

price_prod <- honey_data %>%
  group_by(year) %>%
  summarize(
    totalprod = sum(totalprod, na.rm = TRUE),
    priceperlb = mean(priceperlb, na.rm = TRUE)
  )

p3 <- ggplot(
  price_prod,
  aes(x = totalprod, y = priceperlb, label = year)
) +
  geom_point(alpha = 0.85, size = 3.2, color = honey_gold) +
  geom_text_repel(
    size = 3.2,
    box.padding = 0.35,
    max.overlaps = 20,
    color = honey_dark
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = honey_dark,
    linewidth = 1
  ) +
  scale_x_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    x = "U.S. total production",
    y = "Average price per lb",
    title = str_wrap("Annual U.S. Production Compared with Honey Price", width = 45),
    subtitle = str_wrap("Each point represents one year", width = 65),
    caption = "Data: honeyproduction.csv"
  ) +
  theme_honey_light()

p3
```



## Production Intensity Across Top Honey-Producers

```{r}
#| fig-width: 11
#| fig-height: 6
#| out-width: "100%"
#| fig-align: center

top_states <- honey_data %>%
  group_by(state) %>%
  summarize(mean_totalprod = mean(totalprod, na.rm = TRUE)) %>%
  slice_max(mean_totalprod, n = 15) %>%
  pull(state)

heat_df <- honey_data %>%
  filter(state %in% top_states) %>%
  group_by(state, year) %>%
  summarize(totalprod = mean(totalprod, na.rm = TRUE), .groups = "drop")

p4 <- ggplot(
  heat_df,
  aes(
    x = year,
    y = fct_reorder(state, totalprod, .fun = mean),
    fill = totalprod
  )
) +
  geom_tile(color = "white", linewidth = 0.45) +
  scale_fill_gradient(
    low = "#FFF8DC",
    high = honey_gold,
    labels = label_number(scale_cut = cut_short_scale())
  ) +
  guides(
    fill = guide_colorbar(
      barheight = unit(0.6, "cm"),
      barwidth = unit(8, "cm")
    )
  ) +
  scale_x_continuous(breaks = seq(1998, 2012, by = 2)) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Production",
    title = str_wrap("Production Intensity Across Top Honey-Producers", width = 45),
    subtitle = str_wrap("Darker gold cells indicate higher total production", width = 65),
    caption = "Data: honeyproduction.csv"
  ) +
  theme_honey_light() +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom"
  )

p4
```



## Top 5 Producing States vs. All Other States

```{r}
#| fig-width: 11
#| fig-height: 6
#| out-width: "100%"
#| fig-align: center

top5_states <- honey_data %>%
  group_by(state) %>%
  summarize(mean_totalprod = mean(totalprod, na.rm = TRUE)) %>%
  slice_max(mean_totalprod, n = 5) %>%
  pull(state)

area_df <- honey_data %>%
  mutate(state_group = if_else(state %in% top5_states, state, "Other")) %>%
  group_by(year, state_group) %>%
  summarize(totalprod = sum(totalprod, na.rm = TRUE), .groups = "drop") %>%
  mutate(state_group = fct_relevel(state_group, "Other", after = Inf))

p5 <- ggplot(
  area_df,
  aes(x = year, y = totalprod, fill = state_group)
) +
  geom_area(alpha = 0.92, color = "white", linewidth = 0.15) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_x_continuous(breaks = seq(1998, 2012, by = 2)) +
  scale_fill_manual(
    values = c(
      "Other" = "#FFE9A3",
      "CA" = "#8C510A",
      "FL" = "#BF812D",
      "ND" = "#D88400",
      "SD" = "#F6C344",
      "MT" = "#FFD966"
    )
  ) +
  labs(
    x = NULL,
    y = "Total production (lbs)",
    fill = "State",
    title = str_wrap("Top 5 Producing States vs. All Other States", width = 45),
    subtitle = str_wrap("Stacked area chart showing how major producers contribute to national totals", width = 70),
    caption = "Data: honeyproduction.csv"
  ) +
  theme_honey_light() +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )

p5
```

